Adaption of the Chumbley Score to matching of bullet striation marks

### 2018-02-11 doc/paper.Rmd
minor changes to limitations of chumbley score

As mentioned earlier, subject bias and error rate determination have been a long standing issue in firearm examination \citep{NAS:2009}. This issue of removing subjectivity and making firearm analysis objective  was also raised in the \citet{pcast2016} report. For an objective analysis, scientific principle testability and error rate determination are considered to be defining aspects.  Identification of the uncertainities associated with a method of comparison and coming up with a systemized method of quantification, is therefore important. Associating mathematical probabilites in a systematic manner to these methods are one such way of quantification. 
In a previous study conducted by \citet{aoas} a machine learning based algorithm was developed for same source matching of bullets and error rates were discussed using the database from the Hamby Study \citep{hamby}. The method proposed by \citet{chumbley} (and later improved by \citet{hadler}) also provides a means to determine error rates and claims to reduce subject bias. Therefore, giving us a strong motivation to explore the adaptibility of the Chumbley score methodology to bullets. In this paper, we try to adapt this deterministic algorithm and method developed for toolmarks, to bullets. After this, we consequently discuss about the efforts in doing so, along with the associated error rates. The data used in this paper also belongs to the Hamby Study \citep{hamby}. This gives us a common platform for comparing the performance of the chumbley method on bullets with the already existing method proposed by \citet{aoas} for bullets. As a brief overview, \citet{chumbley}, in their paper, compare two toolmarks with the intention of determining if it comes from the same tool.  They use an empirical based setup to validate their proposed algorithm and quantitative method which calculates a U-statisic for the purpose of classification of toolmarks as matching or non-matching. The data for their study was obtained from 50 sequentially manufactured screwdriver tips, and preselected comparison window sizes were given as inputs to the algorithm. The algorithm then compares the two toolmarks and comes up with a Mann-Whitney U-statistic and an associated p-value to designate them as matches or non-matches. The performance for every 100 comparisons, of the algorithm proposed by \citet{chumbley} and the improvement proposed by \citet{hadler} are listed in the table below.

@@

## Potential limitations of the Chumbley Score adaptation to bullets 
Bullets are much smaller in length, width, are not flat and curved in the cross-sectional topography as opposed to tools like screw driver tips which produced longer and pronounced markings. This means the makings made by barrels in comparison to toolmarks may have a problem in distinctiveness. The majority of Bullet profiles and signatures extracted by procedures mentioned by  \citet{aoas} are almost 1/4 th the size of toolmarks as used by \citet{chumbley} or even smaller. Striations on Bullets are made on their curved surfaces, whereas the algorithm developed by \citet{chumbley} and \citet{hadler} has only been tested for flatter and wider surfaces which have negligible curvature. Therefore, using methods proposed for toolmarks may need adaptation in order to give tangible results for bullets. Moreover, in order to to get flat bullet signatures and remove the curvatures some kind of smoothing needs to be applied as a pre-step which needs further investigation as to whether the level of smoothing does effect the working of the algorithm on Bullets.

Also when in the optimization step, the Window of optimization for bullets will be shorter as the signatures are smaller. The idea is to keep the number of windows of optimization sufficiently large, which means shorter Trace segments (partition of signature or toolmark with length = size of window of optimization) that lets us compare smaller segments of one signature to another. This introduces a problem as, if we go too small in the window of optimization, the unique features of the trace segments are lost and seem similar, while too large sizes vastly reduces the weight of small features that would otherwise uniquely classify a signature and hence identify the region of agreement.

Thus the Window of Optimization has a direct influence on whether we are Falsely rejecting the null when it is true (Type I) or Falsely accepting the null when it is false (Type II) making identification of the optimum size of window of optimization very important. This raises important questions about what parameter settings need to be chosen for bullet land comparison. Something similar is done in \citet{afte-chumbley} for toolmark comparisons of slip-joint pliers where optimum window sizes are determined. There is a need to figure out the best parameter settings which minimizes the errors for unsmoothed markings and pre-processed signatures as determined by \citet{aoas}. An analysis of these error rates and comparison with other methods will help us understand the adaptibility of the chumbley score to bullets.

\hh{figure \ref{fig:rgl} needs to be mentioned somewhere in the write-up}

@@

The Chumbley score algorithm takes input as two vectorized processes $x(t_1)$, $t_1 = 1,2,...T_1$ and $y(t_2)$, $t_2 = 1,2...T_2$ which denote two sets of marks or striae. Here the processes are of the form $z(s)$ which is a spatial process for some $t$, while $z(s_1)$ is the realization of the process in $s_1$. The $t_1$ and $t_2$ can be better understood as equally spaced pixel locations for the two marks under consideration, where a 'pixel' is the resolution of the confocal light microscope. In the case for bullet signatures a pixel corresponds to 0.645 microns. These marks or striae are potentially from two different bullets or two different toolmarks whose source needs to be identified as being same or different. The marks or striae are indexed by the pixel location where $t_1$ is for the first striae referred to as $x$ and $t_2$ is for the second striae which is referred as $y$. $T_1$ and $T_2$ being the respective lengths of the markings, need not be the same but are usually of similar lenghtsThe similarity is then judged by the algorithm on the basis of cross-correlation of a fixed and constant number consecutive pixels say $k$ taken from the two indexed marks $x(t_1)$ and $y(t_2)$ such that in theory $k$ remains smaller that length of the two striae or marks which is the same as $t_1$ and $t_2$. Depending on the what stage of the algorithm we are in, matching of different pixel lengths and locations is done, which at the end effectively compares all possible windows that would guarantee in quantifying the two marks or striae as coming from the same source or not, which also lets us assess the error rates by checking for a large number of cases.

The algorithm works in two phases, namely, an optimization step and a validation step, at the end of which a Mann Whitney U statistic is calculated. A pre-processing step to the algorithm is to choose a coarseness value which  is used as a parameter to the LOWESS smoothing function. The coarseness essentially gives the proportion of points which influence the smooth at each value, which means larger values lead to more smoothness. The LOWESS smoothing is applied to each of two sets of vectorized striae or marks $x(t_1)$ and $y(t_2)$, before proceeding to the algorithm.

@@

Since we are only interested in a non-parameteric U statistic, Hadler et al. proposed a normalization procedure in the Validation step that goes to some extent to address the issue of mutual dependence. Also the same shift and different shift substep was modified to use a deterministic rule for sampling sam-shhift and dofferent shift-samples as opposed to the originally proposed random samples.

The data that is used by \citet{chumbley} is generated by a surface profilometer that gives the height in terms of distance along a linear trace. This is taken perpendicular to the striations present in the toolmark. Two such trace are then compared using the algorithm.

>>>>

As mentioned earlier, subject bias and error rate determination have been a long standing issue in firearm examination \citep{NAS:2009}. This issue of removing subjectivity and making firearm analysis objective  was also raised in the \citet{pcast2016} report. For an objective analysis, scientific principle testability and error rate determination are considered to be defining aspects.  Identification of the uncertainities associated with a method of comparison and coming up with a systemized method of quantification, is therefore important. Associating mathematical probabilites in a systematic manner to these methods are one such way of quantification. 
In a previous study conducted by \citet{aoas} a machine learning based algorithm was developed for same source matching of bullets and error rates were discussed using the database from the Hamby Study \citep{hamby}. The method proposed by \citet{chumbley} (and later improved by \citet{hadler}) also provides a means to determine error rates and claims to reduce subject bias. Therefore, giving us a strong motivation to explore the adaptibility of the Chumbley score methodology to bullets. In this paper, we take this toolmark based methodology and try to apply and adapt it to bullets. After this, we consequently discuss about the efforts in doing so, along with the associated error rates. The data used in this paper also belongs to the Hamby Study \citep{hamby}. This gives us a common platform for comparing the performance of the chumbley method on bullets with the already existing method proposed by \citet{aoas} for bullets. As a brief overview, \citet{chumbley}, in their paper, compare two toolmarks with the intention of determining if it comes from the same tool.  They use an empirical based setup to validate their proposed algorithm and quantitative method which calculates a U-statisic for the purpose of classification of toolmarks as matching or non-matching. The data for their study was obtained from 50 sequentially manufactured screwdriver tips, and preselected comparison window sizes were given as inputs to the algorithm. The algorithm then compares the two toolmarks and comes up with a Mann-Whitney U-statistic and an associated p-value to designate them as matches or non-matches. The performance for every 100 comparisons, of the algorithm proposed by \citet{chumbley} and the improvement proposed by \citet{hadler} are listed in the table below.

@@

## Potential limitations of the Chumbley Score adaptation to bullets
Bullet striations are different from toolmarks. Bullet markings are typically produced by the barrel on the bullets, while toolmarks can be made by tool tips on different surfaces. Tools like screw driver tips as used by \citet{chumbley}, produce longer and pronounced markings. Bullets on the other hand are much smaller in length, width, and curved in the cross-sectional topography. This means the markings made by barrels as opposed to markings made by tools, may have a problem in distinctiveness. The majority of Bullet profiles and signatures of the Hamby study \citep{hamby} used in this paper are extracted by procedures mentioned by  \citet{aoas}. The Ruger pistol barrels used in the  Hamby study have a 9mm caliber. This means that bullets used in these pistols have land markings of almost $(1/6)^{th}$ the size of screwdriver toolmarks used by \citet{chumbley}.  Also, striations on bullets are made on their curved surfaces, whereas the algorithm developed by \citet{chumbley} and \citet{hadler} has only been tested for flatter and wider surfaces which have negligible curvature. Therefore, using methods proposed for toolmarks may need adaptation in order to give tangible results for bullets. Moreover, in order to get flat bullet signatures and remove the curvatures some kind of smoothing needs to be applied as a pre-step wihch needs further investigation as to whether the level of smoothing does effect the working of the algorithm on Bullets.

The Chumbley method and algorithm works on the premise of comparison windows which are supposed to represent small segments of pixels of equal sizes in the two markings that are to be compared in a predetermined fashion. They term them as windows of optimization and validation. The optimization window is selected in the first step called the optimization step where areas/windows of best agreement between the the two markings are found. The window of optimization for bullet striations are shorter as bullet signatures are smaller compared to toolmarks . The idea is to keep the number of windows of optimization sufficiently large. This means we have shorter trace segments that lets us compare smaller segments of one signature to another than those for toolmarks. Here, trace segments are the partitions of marking with the length of the partition equal to the size of window of optimization. This introduces a problem as, if we go too small in the window of optimization, the unique features of the trace segments are lost and seem similar, while too large sizes vastly reduces the weight of small features that would otherwise uniquely classify a signature and hence identify the region of agreement. The problem of distinctiveness in the structure of bullet striations may also lead to potential failure of the algorithm to identify correct windows of agreements or not be able to identify these windows at all. This may lead to incorrect classification of bullets as matches or non-matches.

Thus, the comaparison windows have a direct influence on the false positives and false negatives, making identification of the optimum sizes of these windows very important. This raises important questions about how and what parameter settings need to be chosen for bullet land comparison. Something similar is done in \citet{afte-chumbley} for toolmark comparisons of slip-joint pliers where optimum window sizes are determined. There is a need to figure out the best parameter settings which minimizes the errors for unsmoothed markings or profiles and pre-processed signatures. An analysis of these error rates and comparison with other methods will help us understand the adaptibility of the chumbley score to bullets.

Figure \ref{fig:rgl} shows us how the extracted signature from a bullet land lines up with the image of the land from which it was extracted. We can see also see in the figure how the depth and relative position of the striation markings seen in the image are interpreted as the signature.

@@

The Chumbley score algorithm takes input as two vectorized processes $x(t_1)$, $t_1 = 1,2,...T_1$ and $y(t_2)$, $t_2 = 1,2...T_2$ which denote two sets of marks or striae. Here the processes are of the form $z(s)$ which is a spatial process for some $t$, while $z(s_1)$ is the realization of the process in $s_1$. The $t_1$ and $t_2$ can be better understood as equally spaced pixel locations for the two marks under consideration, where a 'pixel' is the resolution of the confocal light microscope. In the case for bullet signatures extracted from NIST scans of the Hamby study, a pixel corresponds to approximately 1.56 microns. These marks or striae are potentially from two different bullets or two different toolmarks whose source needs to be identified as being same or different. The marks or striae are indexed by the pixel location where $t_1$ is for the first striae referred to as $x$ and $t_2$ is for the second striae which is referred as $y$. $T_1$ and $T_2$ being the respective lengths of the markings, need not be the same but are usually of similar lenghts. The similarity is then judged by the algorithm on the basis of cross-correlation of a fixed and constant number consecutive pixels say $k$ taken from the two indexed marks $x(t_1)$ and $y(t_2)$ such that in theory $k$ remains smaller that length of the two striae or marks which is the same as $t_1$ and $t_2$. Depending on the what stage of the algorithm we are in, matching of different pixel lengths and locations is done, which at the end effectively compares all possible windows that would guarantee in quantifying the two marks or striae as coming from the same source or not, which also lets us assess the error rates by checking for a large number of cases.

The algorithm works in two phases, namely, an optimization step and a validation step, at the end of which a Mann Whitney U statistic is calculated. A pre-processing step to the algorithm is to choose a coarseness value which  is used as a parameter to the lowess smoothing function. The coarseness essentially gives the proportion of points which influence the smooth at each value, which means larger values lead to more smoothness. The lowess smoothing is applied to each of two sets of vectorized striae or marks $x(t_1)$ and $y(t_2)$, before proceeding to the algorithm.

@@

Since we are only interested in a non-parameteric U statistic, \citet{hadler} proposed a normalization procedure in the Validation step that goes to some extent to address the issue of mutual dependence. Also the same shift and different shift substep was modified by \citet{hadler} to use a deterministic rule for sampling same-shift and different shift-samples as opposed to the originally proposed random samples.

<!-- The data that is used by \citet{chumbley} is generated by a surface profilometer that gives the height in terms of distance along a linear trace. This is taken perpendicular to the striations present in the toolmark. Two such trace are then compared using the algorithm. -->
